---
title: "SeedMatchR_noncanonical_searches"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  eval = FALSE
)
```

```{r setup}
library(SeedMatchR)
```

## Running with wobbles and bulges

```{r}
wobbles_bulges_matches = run_wobble_bulges_pipeline(guide.seq = guide.seq,
                                                    res =res,features = annodb$seqs,
                                                    bulges_lim = 1)
  
head(wobbles_bulges_matches)
```


## 3`-compensatory non-canonical matches

The get_compensatory_sequences function is designed to identify 3' compensatory sequences for a given siRNA sequence. It takes several parameters, including the guide sequence (guide.seq), seed sequence (seed), a set of DNA sequences (sequences), and various limits and mismatch allowances. The function returns a data frame containing the identified compensatory regions.

```{r}

comp_matches = get_compensatory_sequences(guide.seq = guide.seq,
                                                seed = get_seed(guide.seq, 'mer7m8')$Target.Seq,
                                                sequences = annodb$seqs,
                                                compensatory_region_length_min = 6,
                                                limit = 100)

head(comp_matches)
```

```{r, fig.height=5, fig.width=5, out.retina=1}
comp.list = unique(comp_matches$TX_ID)

# Gene set 2
background.list = res$gene_id[!(res$gene_id %in% comp.list)]

ecdf.results = deseq_fc_ecdf(res, 
                             list("Background" = background.list, 
                                  "Compensetory Matches" = comp.list))

ecdf.results$plot
```
