---
title: "SeedMatchR output formats"
author: "Tareian Cazares"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SeedMatchR output formats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SeedMatchR)
```

This example uses the siRNA sequence, D1, targeting the Ttr gene in rat liver from the publication:

> Schlegel MK, Janas MM, Jiang Y, Barry JD, Davis W, Agarwal S, Berman D, Brown CR, Castoreno A, LeBlanc S, Liebow A, Mayo T, Milstein S, Nguyen T, Shulga-Morskaya S, Hyde S, Schofield S, Szeto J, Woods LB, Yilmaz VO, Manoharan M, Egli M, Charissé K, Sepp-Lorenzino L, Haslett P, Fitzgerald K, Jadhav V, Maier MA. From bench to bedside: Improving the clinical safety of GalNAc-siRNA conjugates using seed-pairing destabilization. Nucleic Acids Res. 2022 Jul 8;50(12):6656-6670. doi: 10.1093/nar/gkac539. PMID: 35736224; PMCID: PMC9262600.

The guide sequence of interest is 23 bp long and oriented 5' -\> 3'.

```{r}
# siRNA sequence of interest targeting a 23 bp region of the Ttr gene
guide.seq = "UUAUAGAGCAAGAACACUGUUUU"
```

#### Load annotation databases

```{r}
annodb = load_annotations(reference.name = "rnor6", canonical = FALSE, min.feature.width = 8, longest.utr = T)
```

### Prepare DESEQ2 Results

The DESeq2 results are available through the names `Schlegel_2022_Ttr_D1_30mkg`, `Schlegel_2022_Ttr_D4_30mkg` and `Schlegel_2022_Ttr_D1_10mkg`. The data set name is long, so it will be renamed to `res`. The DESeq2 results file is then filtered. The function `filter_res()` can be used to filter a results file by log2FoldChange, padj, baseMean, and remove NA entries.

```{r echo = T, results = 'hide', message=FALSE, warning=FALSE, error=FALSE}
get_example_data("sirna")

sirna.data = load_example_data("sirna")

res <- sirna.data$Schlegel_2022_Ttr_D1_30mkg

res = filter_res(res, fdr_cutoff=1, fc_cutoff=0)
```

### DESeq2 column output

```{r}
res.de = SeedMatchR(res = res, 
                 seqs = annodb$seqs, 
                 sequence = guide.seq, 
                 seed.name = "mer7m8")

head(res.de)
```

### Data frame output

You can perform a seed match for a single seed using the `SeedMatchR()` function. 

```{r}
res.df = SeedMatchR(seqs = annodb$seqs, 
                 sequence = guide.seq, 
                 seed.name = "mer7m8", 
                 res.format = "data.frame")

head(res.df)
```

### GRanges output

You can perform a seed match for a single seed using the `SeedMatchR()` function. 

```{r}
res.gr = SeedMatchR(seqs = annodb$seqs, 
                 sequence = guide.seq, 
                 seed.name = "mer7m8", 
                 res.format = "granges")

head(res.gr)
```

### IRanges output

You can perform a seed match for a single seed using the `SeedMatchR()` function. 

```{r}
res.ir = SeedMatchR(res = res, 
                 seqs = annodb$seqs, 
                 sequence = guide.seq, 
                 seed.name = "mer7m8", 
                 res.format = "iranges")

head(res.ir)
```
