---
title: "SeedMatchR ECDF Statistics"
author: "Tareian Cazares"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SeedMatchR_ECDFStats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SeedMatchR)
```

# Examples of ECDF statisics

There are a variety of approaches that can be used to calculate the difference between two distributions. Historically the KS Dstat has been used for miRNA related studies, but we have also explored other statistics such as the Wasserstein. The differences are outlined in the [TwoSamples package]("https://twosampletest.com/reference/two_sample.html#ref-usage").

## Example workflow

```{r}
# siRNA sequence of interest targeting a 23 bp region of the Ttr gene
guide.seq = "UUAUAGAGCAAGAACACUGUUUU"
```

### Load example DESeq2 data

```{r}
get_example_data("sirna")

sirna.data = load_example_data("sirna")

res <- sirna.data$Schlegel_2022_Ttr_D1_30mkg

res = filter_res(res)
```

### Load Annotations

We will next load the annotations for rat. This step is further described in the vignette.

```{r echo = T, results = 'hide', message=FALSE, warning=FALSE, error=FALSE}
# Load the species specific annotation database object
annodb = load_annotations(reference.name = "rnor6", canonical = FALSE, min.feature.width = 8, longest.utr = T)
```

### Search features for seed matches

```{r}
res = SeedMatchR(res = res, 
                 seqs = annodb$seqs, 
                 sequence = guide.seq, 
                 seed.name = "mer7m8")

head(res, 2)
```

### Extract gene matches for each criteria and their fold changes

```{r}
# Gene set 1 
mer7m8.list = res$gene_id[res$mer7m8 >= 1]

# Gene set 2
background.list = res$gene_id[res$mer7m8 == 0]

# Gene set 1 FC
mer7m8.FC = res$log2FoldChange[res$mer7m8 >= 1]

# Gene set 2 FC
background.FC = res$log2FoldChange[res$mer7m8 == 0]
```

### KS statistic

```{r, fig.height = 5.5, fig.width=5}
ecdf.results = deseq_fc_ecdf(res, 
                             list("Background" = background.list, 
                                  "mer7m8" = mer7m8.list), 
                             stats_test = "KS")

ecdf.results$plot
```

### Wasserstein statistic

```{r, fig.height = 5.5, fig.width=5}
ecdf.results = deseq_fc_ecdf(res, 
                             list("Background" = background.list, 
                                  "mer7m8" = mer7m8.list), 
                             stats_test = "Wass")

ecdf.results$plot
```

### DTS

```{r, fig.height = 5.5, fig.width=5}
ecdf.results = deseq_fc_ecdf(res, 
                             list("Background" = background.list, 
                                  "mer7m8" = mer7m8.list), 
                             stats_test = "DTS"
                             )

ecdf.results$plot

```

### Wilcoxen statistic

```{r, fig.height = 5.5, fig.width=5}
ecdf.results = deseq_fc_ecdf(res, 
                             list("Background" = background.list, 
                                  "mer7m8" = mer7m8.list), 
                             stats_test = "Wilcoxen")

ecdf.results$plot
```
