---
title: "SeedMatchR Search Details"
author: "Tareian Cazares, Pradeep Pugalenthi, and Gulcin Ozer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SeedMatchR_SearchDetails}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  eval = TRUE
)
```

```{r setup}
library(SeedMatchR)
library(Biostrings)
```

## Overview

`SeedMatchR` uses `Biostrings::matchPattern` variants for searching sequences. This article describes different parameters of the search and their results. 

## Load example data

This example uses the siRNA sequence, D1, targeting the Ttr gene in rat liver from the publication:

    Schlegel MK, Janas MM, Jiang Y, Barry JD, Davis W, Agarwal S, Berman D, Brown CR, Castoreno A, LeBlanc S, Liebow A, Mayo T, Milstein S, Nguyen T, Shulga-Morskaya S, Hyde S, Schofield S, Szeto J, Woods LB, Yilmaz VO, Manoharan M, Egli M, Charissé K, Sepp-Lorenzino L, Haslett P, Fitzgerald K, Jadhav V, Maier MA. From bench to bedside: Improving the clinical safety of GalNAc-siRNA conjugates using seed-pairing destabilization. Nucleic Acids Res. 2022 Jul 8;50(12):6656-6670. doi: 10.1093/nar/gkac539. PMID: 35736224; PMCID: PMC9262600.

### Load sequence data

The guide sequence of interest is 23 bp long and oriented 5' -\> 3'.

```{r}
# siRNA sequence of interest targeting a 23 bp region of the Ttr gene
guide.seq = "UUAUAGAGCAAGAACACUGUUUU"

# siRNA target sequence
ttr.target = get_seed(guide.seq = guide.seq, seed.name = "Full")$Target.Seq

ttr.target.wobble = get_seed(guide.seq = guide.seq, seed.name = "Full", allow_wobbles = TRUE)$Target.Seq

```

### Load annotation databases

```{r}
annodb = load_annotations(reference.name = "rnor6", canonical = FALSE, min.feature.width = 8, longest.utr = T)
```

### Load example transcript

We will use `SeedMatchR` to load an example transcript for Ttr in rats. This sequence is the target of an siRNA that is used in this package. 

```{r}
# Original target sequence
ttr.base.target = DNAString("AAAACAGTGTTCTTGCTCTATAA")

# Ttr + 1 insert @ 12
ttr.ins.target = DNAString("AAAACAGTGTTGCTTGCTCTATAA")

# Ttr + 4 mm @ 12-15 + 1 deletion @ 21 + 1 insert @ 4
ttr.mm4.in.del.target = DNAString("AAATACAGTGTTGCCACTCTAAA")

# Ttr + wobble on mRNA
ttr.wobble.mrna = DNAString("AAAACAGUGTTCTTGCTCTATAA")

```

### Load example DE data

```{r}
get_example_data("sirna")

sirna.data = load_example_data("sirna")

res <- sirna.data$Schlegel_2022_Ttr_D1_30mkg

res = filter_res(res, fdr_cutoff=1, fc_cutoff=0)
```

## Search variants

### Base Ttr example

We expect to see a single complete match

```{r}
matchPattern(ttr.target, ttr.base.target, max.mismatch = 0, with.indels = F)
```

### Ttr with insertion

We expect the match not to be found without the proper flags. 

```{r}
matchPattern(ttr.target, ttr.ins.target, max.mismatch = 0, with.indels = F)
```

We expect this search to have no matches despite the indel flag since the max.mismatch argument is at 0. 

```{r}
matchPattern(ttr.target, ttr.ins.target, max.mismatch = 0, with.indels = T)
```

We expect there to be no match since there is no indels allowed, but we still allow for mismatches. 

```{r}
matchPattern(ttr.target, ttr.ins.target, max.mismatch = 1, with.indels = F)
```

We expect the search to find 1 match since max.mismatch is set to 1 and indels are allowed. 

```{r}
matchPattern(ttr.target, ttr.ins.target, max.mismatch = 1, with.indels = T)
```

## Testing multiple mismatches and indels and how they are reported

The total edit distance from the original sequence is 6 (4 mismatches, 1 insertion, 1 deletion). We want to see how these different parameters are counted in the max.mismatch category with indels. 

We expect no matches.

```{r}
matchPattern(ttr.target, ttr.mm4.in.del.target, max.mismatch = 4, with.indels = F)
```

We expect no matches. 

```{r}
matchPattern(ttr.target, ttr.mm4.in.del.target, max.mismatch = 4, with.indels = T)
```

We expect no matches. 

```{r}
matchPattern(ttr.target, ttr.mm4.in.del.target, max.mismatch = 6, with.indels = F)
```

We expect 1 match. 

```{r}
matchPattern(ttr.target, ttr.mm4.in.del.target, max.mismatch = 6, with.indels = T)
```

```{r}
vcountPattern(ttr.target, DNAStringSet(ttr.mm4.in.del.target), max.mismatch = 6, with.indels = T)
```

## Testing how wobbles are reported

```{r}
ttr.target.wobble
```


```{r}
ttr.base.target
```

```{r}
ttr.wobble.mrna
```


```{r}
matchPattern(ttr.target.wobble, ttr.base.target, max.mismatch = 1)
```

